---
title: "Lab 13 - Colonizing Mars"
author: "Sophie Boyd"
date: "3-27-26"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse) 
library(ggplot2)
library(MASS)
```

### Exercise 1: Simulating our colonists

#### 1.1

```{r age}
set.seed(123)
age <- rnorm(100, mean = 30, sd = 5)
```

```{r attributes}
df_colonists <- data.frame(
  id = 1:100,
  age = rnorm(100, mean = 30, sd = 5)
)
```

```{r age-dist}
df_colonists %>%
  ggplot(aes(x = age)) +
  geom_histogram(binwidth=5) +
  labs(x = 'Age', 
       y = 'Frequency') 
```

#### 1.2

Age follows a normal distribution, centered at 30 years old. 
The spread of the distribution is the same regardless of seed because the standard deviation was set at the same value each time (referring to the plot in the lab instructions). 

#### 1.3

```{r role}
df_colonists$role <- rep(c("engineer", "scientist", "medic"),
                         times = c(33, 33, 33),
                         length.out = 100
                         )
```


I wanted approximately equal numbers of engineers, medics, and scientists, so I used method C.

#### 1.4

```{r marsgar}
set.seed(123)

df_colonists$marsgar <- runif(n = 100, min = 0, max = 100)
```

```{r marsgar-by-age, message=FALSE}
df_colonists %>%
  ggplot(aes(x = age, y = marsgar)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

No relationship between age and MARSGAR score due to simulation specifications.

### Exercise 2: Growing our colonists

#### 2.1

```{r tech-skills}
set.seed(123)

df_colonists$technical_skills <- 2 * df_colonists$age + rnorm(100, mean = 0, sd = 1)
```

```{r age-tech-plot, message=FALSE}
df_colonists %>%
  ggplot(aes(x = age, y = technical_skills)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = 'Age',
       y = 'Technical Skills')
```

#### 2.2

```{r prob-solving-role}
set.seed(123)

df_colonists$problem_solving[df_colonists$role == "engineer"] <- rnorm(sum(df_colonists$role == "engineer"), mean = 9, sd = 1)
df_colonists$problem_solving[df_colonists$role == "scientist"] <- rnorm(sum(df_colonists$role == "scientist"), mean = 7, sd = 1)
df_colonists$problem_solving[df_colonists$role == "medic"] <- rnorm(sum(df_colonists$role == "medic"), mean = 6, sd = 1)
```

```{r ps-role-plot}
df_colonists %>%
  ggplot(aes(x = problem_solving, fill = role)) +
  geom_density(alpha = .5) + 
  labs(x = 'Problem-Solving Skills',
       y = 'Count',
       fill = 'Role') +
  scale_fill_viridis_d() +
  theme_minimal()
```

### Exercise 3: Exploring correlations with mvnorm

#### 3.2

```{r agreeableness-resilience, message=FALSE}
set.seed(123)

mean_traits <- c(50, 50)
cov_matrix <- matrix(c(100, 50, 50, 100), ncol = 2)

traits_data <- mvrnorm(n = 100, mu = mean_traits, Sigma = cov_matrix, empirical = FALSE)

colnames(traits_data) <- c("Resilience", "Agreeableness")

traits_dataframe <- as.data.frame(traits_data)

traits_dataframe %>%
  ggplot(aes(x = Resilience, y = Agreeableness)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Resilience", y = "Agreeableness") +
  theme_minimal()

df_colonists <- cbind(df_colonists, traits_dataframe)
```

#### 3.3

```{r bigfive-simulation, message=FALSE}
seed <- 123
set.seed(seed)

library(MASS)
library(Matrix)
library(tidyverse)
library(conflicted)
conflicts_prefer(dplyr::select())

n_colonists <- 100

var_names <- c("EX", "ES", "AG", "CO", "OP")

mean_traits <- c(-.5, .5, .25, .5, 0) 
sd_traits <- c(1, .9, 1, 1, 1) 


cor_matrix_bigfive <- matrix(
  c(
1.0000, 0.2599, 0.1972, 0.1860, 0.2949,
0.2599, 1.0000, 0.1576, 0.2306, 0.0720,
0.1972, 0.1576, 1.0000, 0.2866, 0.1951,
0.1860, 0.2306, 0.2866, 1.0000, 0.1574,
0.2949, 0.0720, 0.1951, 0.1574, 1.0000
  ),
  nrow = 5, ncol = 5, byrow = TRUE,
  dimnames = list(
c("EX", "ES", "AG", "CO", "OP"),
c("EX", "ES", "AG", "CO", "OP")
  )
)

cov_matrix_bigfive <- cor_matrix_bigfive * (sd_traits %*% t(sd_traits))


bigfive_data <- mvrnorm(n = 100, mu = mean_traits, Sigma = cov_matrix_bigfive)

bigfive_data <- cbind.data.frame(
  colonist_id = 1:n_colonists,
  seed = seed, 
  bigfive_data
) 
```


```{r bigfive-summary}

summary_stats_mean <- bigfive_data %>%
  select(-colonist_id, -seed) %>%
  summarize(across(everything(), list(mean = mean))) %>%
  rbind(mean_traits)

summary_stats_sd <- bigfive_data %>%
  select(-colonist_id, -seed) %>%
  summarize(across(everything(), list(sd = sd))) %>%
  rbind(sd_traits) 

summary_stats <- cbind(summary_stats_mean, summary_stats_sd)


summary_stats_cor <- bigfive_data %>%
  select(-colonist_id, -seed) %>%
  cor()  %>%
  rbind(cor_matrix_bigfive)

summary_stats
summary_stats_cor
```

The means, standard deviations, and correlations from the simulation were close to the defined population parameters.

### Exercise 4: Preparing for the unexpected

#### 4.1

##### Running the repeated simulations and extracting descriptives/correlations by rep

```{r bigfive-replication}
library(dplyr)
conflicts_prefer(dplyr::select())
set.seed(123)
num_simulations <- 100

all_simulations <- data.frame()

for (i in 1:num_simulations) {
  simulated_data <- mvrnorm(n = 100, mu = mean_traits, Sigma = cov_matrix_bigfive) 
  
  simulated_data <- cbind.data.frame(
colonist_id = 1:n_colonists, 
rep = i,
simulated_data
  ) 

  all_simulations <- rbind(all_simulations, simulated_data)
}

summary_stats <- all_simulations %>%
  select(EX, rep) %>%
  group_by(rep) %>%
  summarize(across(everything(), list(mean = mean, sd = sd)))

summary_stats

summary_stats_cor <- all_simulations %>%
  select(rep,EX, OP) %>%
  group_by(rep) %>%
  summarise(correlation = cor(EX, OP))

summary_stats_cor
```

##### Plotting the EX-OP correlations across the simulations

```{r all-sim-cor-plot}
summary_stats_cor %>%
  ggplot(aes(x = correlation)) +
  geom_density() + 
  labs(x = 'Extraversion-Openness Correlation Coefficients',
       y = 'Count')
```

The correlations between extraversion and openness across the 100 simulations were consistently positive, but ranged in strength. To get a more stable estimate of the correlation, we would need to increase the simulated sample size past n = 100, maybe up to 250 or 500. 

##### Plotting the means and standard deviations across the simulations

```{r mean-parameter-plot}
summary_stats %>%
  ggplot(aes(x = EX_mean)) +
  geom_density() +
  annotate("text", x = -0.5, y = 4.75, label = paste("Population Mean = -.5")) +
  labs(x = 'Mean of Extraversion',
       y = 'Count')
```

```{r sd-parameter-plot}
summary_stats %>%
  ggplot(aes(x = EX_sd)) +
  geom_density() +
  annotate("text", x = 1, y = 5.75, label = paste("Population SD = 1")) +
  labs(x = 'Standard Deviation of Extraversion',
       y = 'Count')
```

