---
title: "Lab 13 - Colonizing Mars"
author: "Sophie Boyd"
date: "3-27-26"
output: github_document
---

### Load packages and data

Having trouble installing MASS... come back later if needed

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse) 
library(ggplot2)
```

### Exercise 1: Simulating our colonists

#### 1.1

```{r age}
set.seed(123)
age <- rnorm(100, mean = 30, sd = 5)
```

```{r attributes}
df_colonists <- data.frame(
  id = 1:100,
  age = rnorm(100, mean = 30, sd = 5)
)
```

```{r age-dist}
df_colonists %>%
  ggplot(aes(x = age)) +
  geom_histogram(binwidth=5) +
  labs(x = 'Age', 
       y = 'Frequency') 
```

Age follows a normal distribution, centered at around 30 years old. The spread of the distribution is the same regardless of seed because the standard deviation was set at the same value each time (referring to the plot in the lab instructions). 

#### 1.2

```{r role}
set.seed(1)
roleA <- rep(c("engineer", "scientist", "medic"),
  length.out = 100
)
# this works if we want to set a specific number of each role
roleB <- rep(c("engineer", "scientist", "medic"),
  each = 34,
  length.out = 100
)

roleC <- rep(c("engineer", "scientist", "medic"),
  times = c(33, 33, 33),
  length.out = 100
)

# if you want to use sampling weights
roleD <- sample(c("engineer", "scientist", "medic"),
  replace = TRUE,
  size = 100, prob = c(1, 1, 1)
)

# if you want to randomly shuffle

roleE <- sample(roleB, size = 100, replace = FALSE)

# my final code

df_colonists$role <- rep(c("engineer", "scientist", "medic"),
                         times = c(33, 33, 33),
                         length.out = 100
                         )
```

#### 1.3

I wanted about equal numbers of engineers, medics, and scientists, so I used method C.

#### 1.4

```{r marsgar}
set.seed(123)

df_colonists$marsgar <- runif(n = 100, min = 0, max = 100)
```

```{r marsgar-by-age, message=FALSE}
df_colonists %>%
  ggplot(aes(x = age, y = marsgar)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

No relationship between age and MARSGAR score.

### Exercise 2: Growing our colonists

#### 2.1

```{r tech-skills}
set.seed(123)

df_colonists$technical_skills <- 2 * df_colonists$age + rnorm(100, mean = 0, sd = 1)
```

```{r age-tech-plot, message=FALSE}
df_colonists %>%
  ggplot(aes(x = age, y = technical_skills)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = 'Age',
       y = 'Technical Skills')
```

#### 2.2

```{r prob-solving-role}
set.seed(123)

df_colonists$problem_solving[df_colonists$role == "engineer"] <- rnorm(sum(df_colonists$role == "engineer"), mean = 9, sd = 1)
df_colonists$problem_solving[df_colonists$role == "scientist"] <- rnorm(sum(df_colonists$role == "scientist"), mean = 7, sd = 1)
df_colonists$problem_solving[df_colonists$role == "medic"] <- rnorm(sum(df_colonists$role == "medic"), mean = 6, sd = 1)
```

```{r ps-role-plot}
df_colonists %>%
  ggplot(aes(x = problem_solving, fill = role)) +
  geom_density(alpha = .5) + 
  labs(x = 'Problem-Solving Skills',
       y = 'Count',
       fill = 'Role') +
  scale_fill_viridis_d() +
  theme_minimal()
```

### Exercise 3: Exploring correlations with mvnorm


